General purpose gene annotation using deep learning

### 2020-04-07 paper/main.tex
table naming

We pre-processed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Supplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{Tab:01}.

The classes are not balanced equally and the tradeoff between the number of intergenic and intronic bases is quite different in both groups of genomes. Table \ref{Tab:statistics_all} shows the ratio of class occurences, together with other overall metrics.

\begin{table}[!t]
\processtable{Data group statistics for all data\label{Tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{Values are averages of the individual values of each genome in a data group. All statistics except the average gene lengths are excluding FASTA sequences without a gene and subsequences that were masked as completely erroneous. Each strand of DNA was counted seperately. The average gene length was determined by the average length of the longest transcript by cumulative exon length, which is how we query for genes during data generation. Brackets show the standart deviation.}
\end{table}

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation (explanation why this is a sign of error?). As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to effectively blank out those bases during training. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{Tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Section \ref{sec:training_data}.

@@

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

@@

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{Tab:01}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in a one hot encoding\\
	  & for (A, T, C, G), otherwise different, see suplementary .. ()\\
Output & Labels in a 4-dimensional one hot encoding \\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{}
\end{table}

@@

\begin{table}[!t]
\processtable{Experimental Results and Comparison\label{Tab:03}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Augustus & 0.632 (+-0.089) & 0.749 (+-0.161)\\
Dilated CNN Baseline & - (+-) & - (+-)\\
Best Helixer model & - (+-) & - (+-) \\
Best model + varied input length & - (+-) & - \\
Best model (+ varied input length) + overlapping &  0.85 (+-0.068) & 0.843 (+-0.151) \\
(Varied input length +) overlapping + ensembling & 0.868 (+-0.064) & 0.863 (+-0.148) \\
\botrule
\end{tabular}}{Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup. The Augustus result is the median score for all genomes.}
\end{table}

\section{Results}
We evaluate our performance based on the average cross-species Subgenic F1 for all genomes except the few used during training. The results are displayed in Table \ref{Tab:03} and show a clear improvement over the AUGUSTUS and dilated CNN baselines. We also demonstrate the effectiveness of two techniques used during inference with the usage of longer input lengths being enabled by our LSTM stack without any fully connected layers on top. To our knowledge, this technique was not used before with a model used for gene annotation.

@@

\begin{table}[!t]
\processtable{Hyperparameters\label{Tab:02}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Value \\
\midrule
Layers & 4\\
Units per layer & 256\\
Learning rate & 1e-4\\
Optimizer & Adam\\
\botrule
\end{tabular}}{}
\end{table}

@@

\begin{table}[!t]
\processtable{Data group statistics for our training data\label{Tab:statistics_training}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.489 (+- 2.073) & 0.914 (+- 0.934) \\
Average gene length & 25,672 (+- 16,605) & 3,509 (+- 906)\\
Geenuff error rate & 0.253 (+- 0.138) & 0.134 (+- 0.072) \\
Fraction of class Intergenic & 0.752 (+- 0.05) & 0.808 (+- 0.106) \\
Fraction of class UTR & 0.013 (+- 0.011) & 0.033 (+- 0.023) \\
Fraction of class CDS & 0.028 (+- 0.028) & 0.077 (+- 0.052) \\
Fraction of class Intron  & 0.207 (+- 0.023) & 0.083 (+- 0.037) \\
\botrule
\end{tabular}}{The description of Table \ref{Tab:statistics_all} applies here as well.}
\end{table}

>>>>

We pre-processed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Supplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{tab:arrays}.

The classes are not balanced equally and the tradeoff between the number of intergenic and intronic bases is quite different in both groups of genomes. Table \ref{tab:statistics_all} shows the ratio of class occurences, together with other overall metrics.

\begin{table}[!t]
\processtable{Data group statistics for all data\label{tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{Values are averages of the individual values of each genome in a data group. All statistics except the average gene lengths are excluding FASTA sequences without a gene and subsequences that were masked as completely erroneous. Each strand of DNA was counted seperately. The average gene length was determined by the average length of the longest transcript by cumulative exon length, which is how we query for genes during data generation. Brackets show the standart deviation.}
\end{table}

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation (explanation why this is a sign of error?). As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to effectively blank out those bases during training. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Section \ref{sec:training_data}.

@@

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.

We also implemented a dilated CNN baseline as this kind of architecture is widely used when working with DNA data as input. Details of the neural architecture search are given in Section \ref{sec:dcnn} and results are shown in Table \ref{tab:results}.

@@

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{tab:arrays}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in a one hot encoding\\
	  & for (A, T, C, G), otherwise different, see suplementary .. ()\\
Output & Labels in a 4-dimensional one hot encoding \\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{}
\end{table}

@@

\begin{table}[!t]
\processtable{Experimental Results and Comparison\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Augustus & 0.632 (+-0.089) & 0.749 (+-0.161)\\
Dilated CNN Baseline & - (+-) & - (+-)\\
Best Helixer model & - (+-) & - (+-) \\
Best model + varied input length & - (+-) & - \\
Best model (+ varied input length) + overlapping &  0.85 (+-0.068) & 0.843 (+-0.151) \\
(Varied input length +) overlapping + ensembling & 0.868 (+-0.064) & 0.863 (+-0.148) \\
\botrule
\end{tabular}}{Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup. The Augustus result is the median score for all genomes.}
\end{table}

\section{Results}
We evaluate our performance based on the average cross-species Subgenic F1 for all genomes except the few used during training. The results are displayed in Table \ref{tab:results} and show a clear improvement over the AUGUSTUS and dilated CNN baselines. We also demonstrate the effectiveness of two techniques used during inference with the usage of longer input lengths being enabled by our LSTM stack without any fully connected layers on top. To our knowledge, this technique was not used before with a model used for gene annotation.

@@

\begin{table}[!t]
\processtable{Hyperparameters\label{tab:params}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Value \\
\midrule
Layers & 4\\
Units per layer & 256\\
Learning rate & 1e-4\\
Optimizer & Adam\\
\botrule
\end{tabular}}{}
\end{table}

@@

\begin{table}[!t]
\processtable{Data group statistics for our training data\label{tab:statistics_training}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.489 (+- 2.073) & 0.914 (+- 0.934) \\
Average gene length & 25,672 (+- 16,605) & 3,509 (+- 906)\\
Geenuff error rate & 0.253 (+- 0.138) & 0.134 (+- 0.072) \\
Fraction of class Intergenic & 0.752 (+- 0.05) & 0.808 (+- 0.106) \\
Fraction of class UTR & 0.013 (+- 0.011) & 0.033 (+- 0.023) \\
Fraction of class CDS & 0.028 (+- 0.028) & 0.077 (+- 0.052) \\
Fraction of class Intron  & 0.207 (+- 0.023) & 0.083 (+- 0.037) \\
\botrule
\end{tabular}}{The description of Table \ref{tab:statistics_all} applies here as well.}
\end{table}
